---
title: "week-2-data-cleaning"
author: "Joaquin Sandoval"
format: html
editor: visual
---
```{r}
#| message: false  
 
library(tidyverse)
library(janitor)

# Read in snow_cover.csv file exported to data/processed which contains a corrected snow_cover column

snow_fixed <- read_csv(here::here("data", "processed", "snow_cover.csv")) |> 
  clean_names()

# View columns and data types of each column 

glimpse(snow_fixed)

```
As expected, `snow_cover` is a <dbl>. `water_cover`, `land_cover`, and `total_cover` are <chr> type. This needs to be addressed. 

### Cleaning `water_cover` column 

```{r}
#| warning: false  

# Display non-numeric values in water_cover column 

snow_fixed |> 
  count(water_cover) |> 
  filter(is.na(as.numeric(water_cover)))
```
`water_cover` contains "-", ".", "n/a", and "unk" values. This explains why it was <chr> type. 

```{r}
# Individually, convert each non-numeric value in water_cover column to NA values 

water_fixed <- snow_fixed |> 
  mutate(water_cover = ifelse(water_cover == "-", NA, water_cover), 
         water_cover = ifelse(water_cover == ".", NA, water_cover),
         water_cover = ifelse(water_cover == "n/a", NA, water_cover),
         water_cover = ifelse(water_cover == "unk", NA, water_cover))

# Check that only NA values remain 

water_fixed|> 
  count(water_cover) |> 
  filter(is.na(as.numeric(water_cover)))

```

```{r}
# Convert water_cover column to numeric 

water_fixed <- water_fixed |> 
  mutate(water_cover = as.numeric(water_cover), 
         
         # Convert values greater than 100 to NA 
         
          water_cover = ifelse(water_cover > 100, NA, water_cover), 
         
         # Convert values less than 0 to NA (precationary measure)
         
          water_cover = ifelse(water_cover < 0, NA, water_cover))

# Very class of water_cover has been corrected to <dbl>

class(water_fixed$water_cover) 

# Verify values fall between 0 and 100 

summary(water_fixed$water_cover)

```

### 'land_cover` cleaning - repeat same steps as above

```{r}
#| warning: false  

# Display non-numeric values in land_cover colunmn 

water_fixed |> 
  count(land_cover) |> 
  filter(is.na(as.numeric(land_cover)))

```

```{r}
#| warning: false  

# Individually remove non-numeric values 

land_fixed <- water_fixed |> 
  mutate(land_cover = ifelse(land_cover == "-", NA, land_cover), 
         land_cover = ifelse(land_cover == ".", NA, land_cover),
         land_cover = ifelse(land_cover == "n/a", NA, land_cover),
         land_cover = ifelse(land_cover == "unk", NA, land_cover))

# Verify that all non-numeric values were converted to NAs

land_fixed|> 
  count(land_cover) |> 
  filter(is.na(as.numeric(land_cover)))
```

```{r}
# Convert land_cover to numeric and remove values below 0 and above 100 

land_fixed <- land_fixed |> 
  mutate(land_cover = as.numeric(land_cover), 
          land_cover = ifelse(land_cover > 100, NA, land_cover), 
          land_cover = ifelse(land_cover < 0, NA, land_cover))

# Verify class is numeric 

class(land_fixed$land_cover) 

# Verify values between 0 and 100 

summary(land_fixed$land_cover)
```
### `total_cover` cleaning 

```{r}
#| warning: false  

# Display all non-numeric values in total_cover column 

land_fixed |> 
  count(total_cover) |> 
  filter(is.na(as.numeric(total_cover)))
```

```{r}
# Individually remove non-numeric values 

total_fixed <- land_fixed |> 
  mutate(total_cover = ifelse(total_cover == "-", NA, total_cover), 
         total_cover = ifelse(total_cover == ".", NA, total_cover),
         total_cover = ifelse(total_cover == "n/a", NA, total_cover),
         total_cover = ifelse(total_cover == "<row r=\\64\\ spans=\\1:11\\ x14ac:dyDescent=\\0.2\\>", NA, total_cover))

# Verify values were removed 

total_fixed|> 
  count(total_cover) |> 
  filter(is.na(as.numeric(total_cover)))

```

```{r}
# Convert total_cover to numeric 

total_fixed <- total_fixed |>
  mutate(Total_cover = as.numeric(total_cover)) 

# Display unique values; this column should only have 100 or NA

unique(total_fixed$total_cover)
```

All rows should have cover counts that sum to 100. If there are any NA values in one cover column and not the other 2, we can calculate that value by adding the other two values and subtracting that from 100. 

```{r}
# Infer missing values and recompute the total_cover column as needed

total_fixed <- total_fixed |>
  mutate(
    
    # Infer missing cover values where one value across snow, water, land columns is NA
    
    # When NA value for snow_cover and we have values for water_cover and land_cover
    # snow_cover = 100 - water_cover - land_cover
    
    snow_cover  = ifelse(is.na(snow_cover)  & !is.na(water_cover) & !is.na(land_cover),
                         100 - water_cover - land_cover, snow_cover),
    
     # When NA value for water_cover and we have values for snow_cover and land_cover
     # water_cover = 100 - snow_cover - land_cover
    
    water_cover = ifelse(is.na(water_cover) & !is.na(snow_cover)  & !is.na(land_cover),
                         100 - snow_cover  - land_cover, water_cover),
    
    # When NA value for land_cover and we have values for snow_cover and water_cover
    # land_cover = 100 - snow_cover - water_cover
    
    land_cover  = ifelse(is.na(land_cover)  & !is.na(snow_cover)  & !is.na(water_cover),
                         100 - snow_cover  - water_cover, land_cover),
    
    # Recompute total_cover with corrected values across all 3 columns 
    
    total_cover = ifelse(snow_cover + water_cover + land_cover == 100,
                         100, NA))

```

```{r}
# Write the file

write_csv(total_fixed, here::here("data", "processed", "all_cover_fixed_JOAQUINSANDOVAL.csv"))

```

